LUCID: Learning-Enabled Uncertainty-Aware Certification of Stochastic Dynamical Systems

### 2025-07-30 docs/paper/appendix.tex
feat: add configuration

\section{Installation}
Provided \texttt{Python>=3.8} is already present, \pylucid can be installed with the command
\begin{lstlisting}[language=bash,numbers=none,xleftmargin=0em]
pip install pylucid[gui,plot] --index-url \
  https://gitlab.com/api/v4/projects/71977529/packages/pypi/simple
\end{lstlisting}
There are a few limitation some of the dependencies introduce, which may preclude the ability of using certain features on not supported platforms.
Formal verification of the barrier requires the \dreal SMT solver, which can only be installed on Linux and non-ARM macOS.
Support for \highs \gls{lp} is not available on Windows at the moment.
The \gurobi solver must be installed beforehand, and requires a license for commercial use, but is free for academic purposes.

For more details on the installation process, please refer to the online documentation at \url{https://lucidtoolsource.gitlab.io/lucid/md_docs_2Pylucid.html}.

@@

We consider the following system:
\begin{equation*}
    \begin{bmatrix}
        {x}_{t+1}
    \end{bmatrix}
    = \begin{bmatrix}
        0.5 {x}_{t}
    \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_1)$.
Given
\begin{align*}
    &\X = [-1, 1] \\
    &\X_0 = [-0.5, 0.5] \\
    &\X_U = [-1, -0.9] \cup [0.9, 1],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=15$ time steps.
The complete configuration for the linear example benchmark is shown in Listing~\ref{lst:linear}.
\lstinputlisting[language=yaml,caption={Configuration for linear example},captionpos=b,label={lst:linear}]{code/linear.yaml}

@@

We consider the following system:
\begin{equation*}
    \begin{bmatrix}
        {x}_{1, t+1} \\
        {x}_{2, t+1}
    \end{bmatrix}
    = \begin{bmatrix}
        {x}_{2, t} - 1 + e^{-x_{1, t}} \\
        -\sin^2(x_{1, t})
    \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_2)$.
Given
\begin{align*}
    &\X = [ -2, 2 ] \times [ -2 , 2 ] \\
    &\X_0 = \{ [x_1, x_2] : (x_1 + 0.5)^2 + (x_2 - 0.5)^2 \leq 0.4 \} \\
    &\X_U = \{ [x_1, x_2] : (x_1 - 0.7)^2 + (x_2 + 0.7)^2 \leq 0.3 \},
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.
The complete configuration for the \barrII benchmark is shown in Listing~\ref{lst:barrier2}.
\lstinputlisting[language=yaml,caption={Configuration for \barrII},captionpos=b,label={lst:barrier2}]{code/barrier2.yaml}

@@

We consider the following system:
\begin{equation*}
    \begin{bmatrix}
        {x}_{1, t+1} \\
        {x}_{2, t+1}
    \end{bmatrix}
    = \begin{bmatrix}
        {x}_{2, t} \\
        \frac{1}{3} {x}^3_{1, t} - {x}_{1,t} - {x}_{2,t}
    \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_2)$.
Given
\begin{align*}
    &\X = [ -3, 2.5 ] \times [ -2 , 1 ] \\
    &\X_0 = [ 1 , 2 ] \times [ -0.7 , 0.3 ] \cup [ -1.8 , -1.4 ] \times [ -0.1 , 0.1 ] \\
    &\qquad \cup [-1.4, -1.2] \times [-0.5 , 0.1] \\
    &\X_U = [ 0.4 , 0.6 ] \times [ 0.2 , 0.6 ] \cup [ 0.6 , 0.7 ] \times [ 0.2 , 0.4 ],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.
The complete configuration for the \barrIII benchmark is shown in Listing~\ref{lst:barrier3}.
\lstinputlisting[language=yaml,caption={Configuration for \barrIII},captionpos=b,label={lst:barrier3}]{code/barrier3.yaml}

@@

We consider a scenario where an autonomous vehicle (AV) controlled by a \gls{nn} is overtaking another vehicle.
The dynamics of the ego vehicle are given by Dubin's car model \EC{does it need a cit or is it just textbook} by adding the noise vector $w = \begin{bmatrix}w_t^1 & w_t^2 & w_t^3\end{bmatrix}$ where each component is drawn from a zero-mean Gaussian with standard deviation $0.01$, $0.01$, and $0.001$, respectively.
The steering wheel angle is supplied by the \gls{nn} controller and we travel at a fixed velocity $v=1$.
Given
\todo{correct the sets}
\begin{align*}
    &\X = [ -3, 2.5 ] \times [ -2 , 1 ] \\
    &\X_0 = [ 1 , 2 ] \times [ -0.7 , 0.3 ] \cup [ -1.8 , -1.4 ] \times [ -0.1 , 0.1 ] \\
    &\qquad \cup [-1.4, -1.2] \times [-0.5 , 0.1] \\
    &\X_U = [ 0.4 , 0.6 ] \times [ 0.2 , 0.6 ] \cup [ 0.6 , 0.7 ] \times [ 0.2 , 0.4 ],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.

>>>>

\section{Tool technical details}

\subsection{Installation}
\OS{Ernesto, can we organise this in a better order?}
The recommended way to install \pylucid is via the \texttt{pip} package manager, using one of the pre-built wheels, 
hence avoiding the time-consuming need to compile the code from source.
Provided \texttt{Python>=3.8} is already present, installing \pylucid requires only the command
\begin{lstlisting}[language=bash,numbers=none,xleftmargin=0em]
pip install pylucid[gui,plot] --index-url \
  https://gitlab.com/api/v4/projects/71977529/packages/pypi/simple
\end{lstlisting}
If the desired platform's wheel is not available, \pylucid can be installed from source by cloning the repository and running
\begin{lstlisting}[language=bash,numbers=none,xleftmargin=0em]
pip install .[gui,plot]
\end{lstlisting}
Note that this requires \texttt{Python>=3.8} and \texttt{Bazel>=8.0} to be installed on the system.
For more details on the installation process, please refer to the online documentation at \url{https://lucidtoolsource.gitlab.io/lucid/md_docs_2Pylucid.html}.

\subsection{Platform support}
Some dependencies \pylucid includes do not support all platforms, which may preclude the ability of using certain features.
Formal verification of the barrier requires the \dreal SMT solver, which can be installed with \lstinline|pip install dreal|, but only supports Linux and non-ARM macOS.
The \highs \gls{lp} solver cannot be used on Windows at the moment.
The \gurobi solver must be installed separately beforehand and requires a licence.

\subsection{Configuration formats}
\pylucid provides extensive support for configuration, which can be specified as a \texttt{.yaml} file (recommended), \texttt{.json} file,
with a \texttt{.py} file returning a \texttt{Configuration} object, or directly via command line arguments.
Moreover, sampled transitions can be embedded in the configuration file, or be provided separately with \texttt{.csv}, \texttt{.mat}, \texttt{.npy}, or \texttt{.npz} file.
For example, following configurations will be equivalent:
\lstinputlisting[language=yaml,caption={Yaml configuration.},captionpos=b,label={lst:configuration.yaml}]{code/configuration.yaml}
\lstinputlisting[language=json,caption={Json configuration.},captionpos=b,label={lst:configuration.json}]{code/configuration.json}
\lstinputlisting[language=iPython,style={nonumbers},caption={Python configuration.},captionpos=b,label={lst:configuration.py}]{code/configuration.py}
\lstinputlisting[language=bash,style={nonumbers},caption={Command line configuration. Assumes that the transition samples have been stored as \texttt{.csv} files.},captionpos=b,label={lst:configuration.sh}]{code/configuration.sh}

\subsection{Graphical User Interface}

\pylucid provides a \gls{gui} to guide the user in the creation of the scenario configuration and presenting the results in an intuitive way.
Running \lstinline|pylucid-gui| will open a browser tab containing the interface shown in Figures~[\ref{fig:gui-left},\ref{fig:gui-right}], while a local server will listen for requests coming from the \gls{gui}, computing and returning the results.
\begin{figure}
    \centering
    \includegraphics[trim={0 12.4cm 15cm 0},clip,width=.7\linewidth]{figures/frontend.png}
    \caption{\lucid's \gls{gui} can be accessed via any web browser. 
    The left half of the screen allows the user to provide the transition samples and the safety specification to verify.
    }
    \label{fig:gui-left}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[trim={0 12.4cm 15cm 0},clip,width=.7\linewidth]{figures/frontend.png}
    \caption{\lucid's \gls{gui} can be accessed via any web browser. 
    The right half of the screen shows the results produced by \lucid, including a plot of the computed barrier function, 
    the numerical safety guarantee and , and detailed logs of the computations.
    }
    \label{fig:gui-right}
\end{figure}

@@

We consider the following system:
\begin{equation*}
      \begin{bmatrix}
            {x}_{t+1}
      \end{bmatrix}
      = \begin{bmatrix}
            0.5 {x}_{t}
      \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_1)$.
Given
\begin{align*}
       & \X = [-1, 1]                     \\
       & \X_0 = [-0.5, 0.5]               \\
       & \X_U = [-1, -0.9] \cup [0.9, 1],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=15$ time steps.
The complete configuration for the linear example benchmark is shown in Listing~\ref{lst:linear}.
\lstinputlisting[language=yaml,caption={Configuration for linear example},captionpos=b,label={lst:linear}]{code/linear.yaml}

@@

We consider the following system:
\begin{equation*}
      \begin{bmatrix}
            {x}_{1, t+1} \\
            {x}_{2, t+1}
      \end{bmatrix}
      = \begin{bmatrix}
            {x}_{2, t} - 1 + e^{-x_{1, t}} \\
            -\sin^2(x_{1, t})
      \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_2)$.
Given
\begin{align*}
       & \X = [ -2, 2 ] \times [ -2 , 2 ]                                  \\
       & \X_0 = \{ [x_1, x_2] : (x_1 + 0.5)^2 + (x_2 - 0.5)^2 \leq 0.4 \}  \\
       & \X_U = \{ [x_1, x_2] : (x_1 - 0.7)^2 + (x_2 + 0.7)^2 \leq 0.3 \},
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.
The complete configuration for the \barrII benchmark is shown in Listing~\ref{lst:barrier2}.
\lstinputlisting[language=yaml,caption={Configuration for \barrII},captionpos=b,label={lst:barrier2}]{code/barrier2.yaml}

@@

We consider the following system:
\begin{equation*}
      \begin{bmatrix}
            {x}_{1, t+1} \\
            {x}_{2, t+1}
      \end{bmatrix}
      = \begin{bmatrix}
            {x}_{2, t} \\
            \frac{1}{3} {x}^3_{1, t} - {x}_{1,t} - {x}_{2,t}
      \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_2)$.
Given
\begin{align*}
       & \X = [ -3, 2.5 ] \times [ -2 , 1 ]                                                 \\
       & \X_0 = [ 1 , 2 ] \times [ -0.7 , 0.3 ] \cup [ -1.8 , -1.4 ] \times [ -0.1 , 0.1 ]  \\
       & \qquad \cup [-1.4, -1.2] \times [-0.5 , 0.1]                                       \\
       & \X_U = [ 0.4 , 0.6 ] \times [ 0.2 , 0.6 ] \cup [ 0.6 , 0.7 ] \times [ 0.2 , 0.4 ],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.
The complete configuration for the \barrIII benchmark is shown in Listing~\ref{lst:barrier3}.
\lstinputlisting[language=yaml,caption={Configuration for \barrIII},captionpos=b,label={lst:barrier3}]{code/barrier3.yaml}

@@

We consider a scenario where an autonomous vehicle (AV) controlled by a \gls{nn} is overtaking another vehicle.
The dynamics of the ego vehicle are given by Dubin's car model \EC{does it need a cit or is it just textbook} by adding the noise vector $w = \begin{bmatrix}w_t^1 & w_t^2 & w_t^3\end{bmatrix}$ where each component is drawn from a zero-mean Gaussian with standard deviation $0.01$, $0.01$, and $0.001$, respectively.
The steering wheel angle is supplied by the \gls{nn} controller and we travel at a fixed velocity $v=1$.
Given
\begin{align*}
       & \X = [ 1, 90 ] \times [ -7, 19 ] \times [ -\pi, \pi ]         \\
       & \X_0 = [ 1, 3 ] \times [ -1, 1 ] \times [ -0.5, 0.5 ]         \\
       & \X_U = [ 1, 90 ] \times [ -7, -6 ] \times [ -\pi, \pi ]       \\
       & \qquad \cup [ 1, 90 ] \times [ 18, 19 ] \times [ -\pi, \pi ]  \\
       & \qquad \cup [ 40, 45 ] \times [ -6, 6 ] \times [ -\pi, \pi ],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.

### 2025-07-30 docs/paper/main.tex
feat: add configuration

We also provide a Python wrapper, called \pylucid, to facilitate the integration of the tool into existing workflows and effortlessly leverage well-established libraries such as NumPy~\cite{numpy} and SciPy~\cite{scipy}.
Moreover, \pylucid can be extensively configured in a variety of ways (e.g., Python scripts, YAML files, GUI), making it the recommended way to operate \lucid.
Hence, for the rest of the paper, we will focus on \pylucid when describing the user interfaces.
{\color{red}Instructions for installing \lucid are deferred to the appendix.}

@@

\paragraph{Dataset}
% \todo{Should this be considered a component? It is more about the data loader} \Sadegh{This is good to have as a separate component.}
Due to its data-driven nature, all that \lucid needs to operate, is a set of sampled transitions $\data_N$ from the system, as shown in \eqref{eq:data}, and the safety specification itself.
% Data can be obtained by defining a function that takes as input the current state and control input and returns the next state, adding i.i.d. noise sampled from an appropriate distribution, or it may be collected from an external simulator or real-world system.
The samples can be specified directly in the configuration file:
\begin{lstlisting}[language=yaml,backgroundcolor=\color{ipython_bg}]
x_samples: [[5.930e-01], ..., [-1.937e-01]]
xp_samples: [[3.015e-01], ..., [-1.018e-01]]
\end{lstlisting}
Alternatively, the samples can either be generated programmatically in the user-provided Python configuration script, or be stored in a separate file.
\pylucid natively supports common formats such as \texttt{.csv}, \texttt{.mat}, \texttt{.npy}, and \texttt{.npz}.
% Refer to the documentation for more details.
\begin{comment}
\begin{lstlisting}[language=iPython]
    def scenario_config():
    # ...
    # Sampling
    f = lambda x: x / 2
    c.x_samples = c.X_bounds.sample(1000)
    c.xp_samples = f(c.x_samples) + np.random.normal(scale=0.01)
    # ...
    return c
\end{lstlisting}
\end{comment}

@@

\paragraph{\Circled{3}~Tuner}
Being parameter-free approaches, kernel methods do not require the expensive learning processes of other machine learning methods, such as neural networks.
However, they still depend on a number of \hp, such as the kernel bandwidth $\sigma_f$, the lengthscale $\sigma_l$, and the regularization constant $\lambda$.
Changes in their values can have a significant impact on the \estimator's efficiency and accuracy.
The process of finding good values for these \hp is known as \emph{hyperparameter tuning}, with the optimal parameters being problem dependent.
\lucid provides a set of utilities, which specialize the \texttt{Tuner} interface, to aid in this task:
\begin{itemize}
    \item \texttt{MedianHeuristicTuner} uses the \emph{median heuristic} \cite{garreau2018largesampleanalysismedian} to produce rule-of-thumb-type estimates for the hyperparameters $\sigma_f$ and $\sigma_l$ of, e.g., the Gaussian kernel \eqref{eq:gaussian_kernel}, via closed-form expressions.
          This can offer a cheap starting point for successive fine-tuning.
    \item \texttt{LbfgsTuner} finds the \hp that maximize the \emph{log marginal likelihood}~\cite{rasmussen2006gaussian}, defined as
          \begin{equation*}
              \begin{split}
                   & \log p(X^+_N | X_N,\theta) =  - \frac{1}{2}X^+_N{}\T (K_{X}^N + N\lambda I_N)^{-1} X^+_N \\
                   & \hspace{40pt} -\frac{1}{2}\log |K_{X}^N + N\lambda I_N| - \frac{N}{2}\log(2\pi),
              \end{split}
          \end{equation*}
          where $\theta:=(\sigma_f,\sigma_l,\lambda)$ is the hyperparameter being optimized.
          The \texttt{LbfgsTuner} uses the L-BFGS or L-BFGS-B quasi-Newton optimization algorithms~\cite{book:nonlinear-programming},
          implemented in the \texttt{LBFGS++} library\footnote{https://github.com/yixuan/LBFGSpp}.
    \item \texttt{GridSearchTuner} implements the grid search method, exploring the space of possible \hp values to find the one that maximizes the \estimator's $R^2$ score, defined as $R^2 = 1 - (\sum_{i=1}^M (y_i - \hat{y}_i)^2/\sum_{i=1}^M (y_i - \bar{y})^2)$.
\end{itemize}
Tuners open a wide range of possibilities to the user.
We recommend using them within a Python script generating a configuration to automate the tuning process before finalizing the \estimator's hyperparameter values:
\begin{lstlisting}[language=iPython,style=nonumbers]
def scenario_config(c: Configuration):
    ...
    t = LbfgsTuner(lb=[1e-5], ub=[1e5])
    c.estimator = KernelRidgeRegressor(
        kernel=GaussianKernel(),
    )
    c.estimator.fit(c.x_samples, c.xp_samples, tuner=t)
    return c
\end{lstlisting}
For the running example, we set the hyperparameters explicitly:
\begin{lstlisting}[language=yaml,backgroundcolor=\color{ipython_bg}]
sigma_f: 15
sigma_l: [1.76]
lambda: 1.0e-5
\end{lstlisting}

@@

\paragraph{\Circled{4}~Feature Map}
We exploit the truncated spectral kernel expansion in \eqref{eq:sqexp_kernel_fourier} by building an explicit approximated feature map, composed of a linear sum of trigonometric functions with increasing frequencies (see Figure~\ref{fig:spectral_barrier}).
The underlying kernel expansion, visualized in Figure~\ref{fig:spectral_measure_abstraction}, can be controlled to trade-off efficiency and accuracy/conserativeness.
% The full expansion of the kernel would need a sum of infinitely many terms with diminishing impact on the overall result, which we truncate arbitrarily short, a decision based on the tradeoff we are looking for between efficiency and accuracy.
After being defined with the same \hp $(\sigma_f,\sigma_l)$ that characterize the \texttt{KernelRidgeRegressor}, the function can be used to map any point from $\X$ to the \gls{rkhs} associated with the kernel.
\lucid currently implements three different ways for partitioning the spectral measure in Figure~\ref{fig:spectral_measure_abstraction}:
\begin{itemize}
    \item \texttt{LinearTruncatedFourierFeatureMap} defines equally spaced frequency bands on the interval $[-3\sigma_l,3\sigma_l]$, capturing 99.73\% of the spectral measure. This partitioning is shown in Figure~\ref{fig:spectral_measure_abstraction}.
    \item \texttt{LogTruncatedFourierFeatureMap} defines logarithmically spaced frequency bands on the interval $[-3\sigma_l,3\sigma_l]$, with each frequency band carrying an equal share of the spectral measure.
    \item \texttt{ConstantTruncatedFourierFeatureMap} defines frequency bands with a fixed size.
\end{itemize}
Different feature map implementations may yield different results.
In our example, we truncate the Fourier expansion to $5$ frequency bands, including the constant term, and use $64$ times the Nyquist frequency of lattice points to build the constraints of the optimization problem, to avoid it being too conservative.\OS{Specifying it as a multiple of the Nyquist frequency will inevitably raise questions. Just say we generate $X$ lattice points? "lattice density"}
\begin{lstlisting}[language=yaml,backgroundcolor=\color{ipython_bg}]
feature_map: 
    "LinearTruncatedFourierFeatureMap"
num_frequencies: 5
oversample_factor: 64
\end{lstlisting}

@@

\paragraph{Validation (optional)}
Albeit, if \lucid finds a barrier, this is already formally sound with respect to the data-driven estimator,
if the system dynamics are known, \lucid provides the option to formally verify the correctness of the barrier using the \dreal SMT solver~\cite{dreal}.
Note that \dreal has to be installed separately with \lstinline|pip install dreal| and may not be available on all systems.

@@

To ascertain the performance of \lucid, we conducted a series of experiments, run on a Windows 10 machine with an AMD Ryzen 9 5950X 16-Core Processor @ 3.40 GHz, and 64 GB of RAM.
All runs had the random seed set to $42$ to ensure reproducibility.

@@

Overview of the experimental results in Table~\ref{tbl:benchmarks_results}.
\begin{table}[tb]
    \centering
    \begin{tabular}{ccccc}
        \toprule
        \textbf{Benchmark} & \textbf{Dim} & \textbf{Lattice} & \textbf{Runtime} & \textbf{Safety Prob.} \\
                           &              & \textbf{Density} & [mm:ss]          & [\%]                  \\ % Units
        \midrule
        Linear             & 1            & $?$              & 00:02            & 95.15                 \\
        Barr3              & 2            & $30^2$           & XX:XX            & XX                    \\
                           &              & $40^2$           & XX:XX            & XX                    \\
                           &              & $?^2$            & XX:XX            & XX                    \\
        Overtaking         & 3            & $?^3$            & XX:XX            & XX                    \\
                           & 3            & $?^3$            & XX:XX            & XX                    \\
                           & 3            & $?^3$            & XX:XX            & XX                    \\
        name2              & 7            &                  & XX:XX            & XX                    \\
        \bottomrule
    \end{tabular}
    \caption{Computational benchmarks: ...}
    \label{tbl:benchmarks_results}
\end{table}

>>>>

We also provide a Python wrapper, called \pylucid, to facilitate the integration of the tool into existing workflows and effortlessly leverage well-established libraries such as NumPy~\cite{numpy} and SciPy~\cite{scipy}.
Moreover, \pylucid can be extensively configured in a variety of ways (e.g., Python scripts, YAML files, GUI), making it the recommended way to operate \lucid.
Hence, for the rest of the paper, we will focus on \pylucid when describing the user interfaces.
{\color{red}Further information about \pylucid are deferred to the appendix.}

@@

\paragraph{Dataset}
% \todo{Should this be considered a component? It is more about the data loader} \Sadegh{This is good to have as a separate component.}
Due to its data-driven nature, all that \lucid needs to operate, is a set of sampled transitions $\data_N$ from the system, as shown in \eqref{eq:data}, and the safety specification itself.
% Data can be obtained by defining a function that takes as input the current state and control input and returns the next state, adding i.i.d. noise sampled from an appropriate distribution, or it may be collected from an external simulator or real-world system.
The samples can be specified directly in the configuration file, divided into $x^1,\ldots,x^N$ and $x^1_+,\ldots,x^N_+$:
\begin{lstlisting}[language=yaml,backgroundcolor=\color{ipython_bg}]
x_samples: [[5.930e-01], ..., [-1.937e-01]]
xp_samples: [[3.015e-01], ..., [-1.018e-01]]
\end{lstlisting}
Alternatively, the samples can either be generated programmatically in the user-provided Python configuration script, or be stored in a separate file.
% Refer to the documentation for more details.
\begin{comment}
\begin{lstlisting}[language=iPython]
    def scenario_config():
    # ...
    # Sampling
    f = lambda x: x / 2
    c.x_samples = c.X_bounds.sample(1000)
    c.xp_samples = f(c.x_samples) + np.random.normal(scale=0.01)
    # ...
    return c
\end{lstlisting}
\end{comment}

@@

\paragraph{\Circled{3}~Tuner}
Being parameter-free approaches, kernel methods do not require the expensive learning processes of other machine learning methods, such as neural networks.
However, they still depend on a number of \hp, such as the kernel bandwidth $\sigma_f$, the lengthscale $\sigma_l$, and the regularization constant $\lambda$ (see \eqref{eq:gaussian_kernel}--\eqref{eq:empiricalEstimate}).
Changes in their values can have a significant impact on the \estimator's efficiency and accuracy.
The process of finding good values for these \hp is known as \emph{hyperparameter tuning}, with the optimal parameters being problem dependent.
\lucid provides a set of utilities, which specialize the \texttt{Tuner} interface, to aid in this task:
\begin{itemize}
    \item \texttt{MedianHeuristicTuner} uses the \emph{median heuristic} \cite{garreau2018largesampleanalysismedian} to produce rule-of-thumb-type estimates for the hyperparameters $\sigma_f$ and $\sigma_l$ of, e.g., the Gaussian kernel \eqref{eq:gaussian_kernel}, via closed-form expressions.
          This can offer a cheap starting point for successive fine-tuning.
    \item \texttt{LbfgsTuner} finds the \hp that maximize the \emph{log marginal likelihood}~\cite{rasmussen2006gaussian}, defined as
          \begin{equation*}
              \begin{split}
                   & \log p(X^+_N | X_N,\theta) =  - \frac{1}{2}X^+_N{}\T (K_{X}^N + N\lambda I_N)^{-1} X^+_N \\
                   & \hspace{40pt} -\frac{1}{2}\log |K_{X}^N + N\lambda I_N| - \frac{N}{2}\log(2\pi),
              \end{split}
          \end{equation*}
          where $\theta:=(\sigma_f,\sigma_l,\lambda)$ is the hyperparameter being optimized.
          The \texttt{LbfgsTuner} uses the L-BFGS or L-BFGS-B quasi-Newton optimization algorithms~\cite{book:nonlinear-programming},
          implemented in the \texttt{LBFGS++} library\footnote{https://github.com/yixuan/LBFGSpp}.
    \item \texttt{GridSearchTuner} implements the grid search method, exploring the space of possible \hp values to find the one that maximizes the \estimator's $R^2$ score, defined as $R^2 = 1 - (\sum_{i=1}^M (y_i - \hat{y}_i)^2/\sum_{i=1}^M (y_i - \bar{y})^2)$.
\end{itemize}
Tuners open a wide range of possibilities to the user.
We recommend using them within a Python script generating a configuration to automate the tuning process before finalizing the \estimator's hyperparameter values:
\begin{lstlisting}[language=iPython,style=nonumbers]
def scenario_config(c: Configuration):
    ...
    t = LbfgsTuner(lb=[1e-5], ub=[1e5])
    c.estimator = KernelRidgeRegressor(
        kernel=GaussianKernel(),
    )
    c.estimator.fit(c.x_samples, c.xp_samples, tuner=t)
    return c
\end{lstlisting}
For the running example, we set the hyperparameters explicitly:
\begin{lstlisting}[language=yaml,backgroundcolor=\color{ipython_bg}]
sigma_f: 15
sigma_l: [1.76]
lambda: 1.0e-5
\end{lstlisting}

@@

\paragraph{\Circled{4}~Feature Map}
We exploit the truncated spectral kernel expansion in \eqref{eq:sqexp_kernel_fourier} by building an explicit approximated feature map, composed of a linear sum of trigonometric functions with increasing frequencies (see Figure~\ref{fig:spectral_barrier}).
The underlying kernel expansion, visualized in Figure~\ref{fig:spectral_measure_abstraction}, can be controlled to trade-off efficiency and accuracy/conserativeness.
% The full expansion of the kernel would need a sum of infinitely many terms with diminishing impact on the overall result, which we truncate arbitrarily short, a decision based on the tradeoff we are looking for between efficiency and accuracy.
After being defined with the same \hp $(\sigma_f,\sigma_l)$ that characterize the \texttt{KernelRidgeRegressor}, the function can be used to map any point from $\X$ to the \gls{rkhs} associated with the kernel.
\lucid currently implements three different ways for partitioning the spectral measure in Figure~\ref{fig:spectral_measure_abstraction}:
\begin{itemize}
    \item \texttt{LinearTruncatedFourierFeatureMap} defines equally spaced frequency bands on the interval $[-3\sigma_l,3\sigma_l]$, capturing 99.73\% of the spectral measure. This partitioning is shown in Figure~\ref{fig:spectral_measure_abstraction}.
    \item \texttt{LogTruncatedFourierFeatureMap} defines logarithmically spaced frequency bands on the interval $[-3\sigma_l,3\sigma_l]$, with each frequency band carrying an equal share of the spectral measure.
    \item \texttt{ConstantTruncatedFourierFeatureMap} defines frequency bands with a fixed size.
\end{itemize}
Different feature map implementations may yield different results.
In our example, we truncate the Fourier expansion to $5$ frequency bands, including the constant term, and use $64$ times the Nyquist frequency of lattice points to build the constraints of the optimization problem, to avoid it being too conservative.\OS{Specifying it as a multiple of the Nyquist frequency will inevitably raise questions. Just say we generate $X$ lattice points? "lattice size"}
\begin{lstlisting}[language=yaml,backgroundcolor=\color{ipython_bg}]
feature_map: 
    "LinearTruncatedFourierFeatureMap"
num_frequencies: 5
oversample_factor: 64
\end{lstlisting}

@@

\paragraph{Validation (optional)}
Albeit, if \lucid finds a barrier, this is already formally sound with respect to the data-driven estimator,
if the system dynamics are known, \lucid provides the option to formally verify the correctness of the barrier using the \dreal SMT solver~\cite{dreal}.

@@

To ascertain the performance of \lucid, we conducted a series of experiments, run on a Windows 10 machine with an AMD Ryzen 9 5950X 16-Core Processor @ 3.40 GHz, and 64 GB of RAM.
All runs had the random seed set to $42$ to ensure reproducibility.
The results of all experimental results are summarized in Table~\ref{tbl:benchmarks_results}.
\begin{table}[tb]
    \centering
    \begin{tabular}{ccccc}
        \toprule
        \textbf{Benchmark} & \textbf{Dim} & \textbf{Lattice} & \textbf{Runtime} & \textbf{Safety} \\
                           &              & \textbf{Size} & [mm:ss]          & \textbf{Probability}                  \\ % Units
        \midrule
        Linear             & 1            & $?$              & 00:02            & 95.15\%                 \\
        Barr2              & 2            & $350^2$           & 14:42            & 75.34\%                    \\
        Barr3              & 2            & $30^2$           & XX:XX            & XX\%                    \\
                           &              & $40^2$           & XX:XX            & XX\%                    \\
                           &              & $?^2$            & XX:XX            & XX\%                    \\
        Overtaking         & 3            & $?^3$            & XX:XX            & XX\%                    \\
                           & 3            & $?^3$            & XX:XX            & XX\%                    \\
                           & 3            & $?^3$            & XX:XX            & XX\%                    \\
        name2              & 7            &                  & XX:XX            & XX\%                    \\
        \bottomrule
    \end{tabular}
    \caption{Computational benchmarks: ...}
    \label{tbl:benchmarks_results}
\end{table}

@@
